ESRC SDAI Application - Understanding domestic abuse using Big Data

### 2019-07-10 case_for_support_TM_hacked.tex
TM comments and hack

 \textcolor{red}{\textit{Short intro about aims, highlighting added value.}} In this project, we will explore the victim and offender characteristics that predict serious harm and the factors determining the decision to report the abuse. Identifying high-risk victim-offender pairs is key in preventing serious harm. Understanding what affects the decision to report is particularly important, as underreporting represents the biggest obstacle to effectively tackling domestic abuse. Insights from these analyses can inform decisions about the optimal timing and target audience for domestic abuse awareness campaigns.
%
%\textcolor{red}{\textit{What we know so far}}
%
%\href{https://www.thehotline.org/2018/09/28/escalation/}{Escalation} often happens when the abuser feels like they are losing control (e.g., because the survivor threatens to leave); separation is an important risk factors of serious harm; contact over children
%
%\href{http://www.safelives.org.uk/policy-evidence/about-domestic-abuse/how-long-do-people-live-domestic-abuse-and-when-do-they-get}{Reporting} 
%
%decision to report: serious violence; friends, family, neighbours notice; happens in public, in front of people

\textcolor{red}{\textit{Detailed plan}} Drawing on the different strengths of the CSEW and the WMP datasets, we will explore the relative importance of the risk factors predicting serious harm. Previous studies investigating this question mostly relied on police data and focused on identifying high-risk offenders \cite{thornton}. Using the CSEW, we will complement and extend this research by simultaneously considering the victim characteristics that are predictive of suffering serious injury. Specifically, using a multinomial logistic approach, we will explore the extent to which financial independence and resilience (income bracket, employment stability), mental health (self-reported well-being), drug and alcohol dependence (frequency of usage), social isolation (self-reported frequency of going out, living alone or with the perpetrator, family disputes), recent separation, and feeling frightened (increased home security, reporting feeling scared) are predictors of various levels of harm (threats, minor or serious injury). We will know the level of harm suffered, because the CSEW asks whether the respondent needed to see a doctor, nurse or other health worker, or needed to take time off work because of the abuse. While we cannot investigate causal relationships due to the cross-sectional nature of the CSEW, this analysis will nevertheless provide us with important insights about the characteristics of groups most vulnerable to domestic abuse.

Using the WMP data, we will also explore the risk factors associated with the perpetrator. The strength of the WMP data is that it has a temporal dimension, and we can follow the same perpetrator-victim pair over time. Our detailed crime dataset will allow to investigate the extent to which the perpetrator's previous recorded self-harm and suicide attempts, previous violent and drug offences, breaches of court orders and separation from the victim will predict causing serious harm, using each perpetrator-victim pair as a unit of analysis. From the detailed description of the incident, we will know if the perpetrator and the victim live together. We will explore this question using a multinomial logistic regression analysis with perpetrator-victim fixed effects, where we will use previously known information about the perpetrator to predict serious harm. Using a fixed-effects regression approach will allow us to disentangle the relative importance of our explanatory variables and perpetrator-victim specific factors in predicting serious harm.

@@

We will also use both datasets to explore the predictors of the decision to report and escape the abuse. We are not aware of previous research exploring this question. Using the CSEW and logistic regression approach, we will investigate how the decision to report (or the decision to seek any form of external help) depends on the individual-level, time-invariant characteristics of the survivor, controlling for the severity and type of the abuse suffered. We are especially interested in exploring how trust in the CJS, social ties (living with family, going out, length of living in the area, member of Neighbourhood Watch), financial independence (own income, employment status, car ownership), number of children, mental health, and alcohol and drug dependency affect the decision to report. Using the WMP data, we will explore time-varying factors affecting the decision to leave the abuser. Due to the high levels of underreporting, it is hard to distinguish between time periods when the number of domestic abuse incidents increase, and when the victims have a higher propensity to finally report the ongoing abuse. Here we are mainly interested in the latter case (and will explore the first case in more detail in the last research project), therefore we will focus on the subset of incidents where the victim for the first time reported a long history of abuse. We are interested in whether these reports are more likely to occur after certain days of the year (e.g., birthday of the victim or perpetrator, Christmas, Halloween, Easter, etc.), potentially reflecting cases where the abuse has suddenly escalated, or became known to outsiders.

@@

 \textcolor{red}{\textit{Short intro about aims, highlighting added value.}}
Domestic abuse has long-lasting adverse effects on victims and those close to them. In this research project, our is aim to explore some of these consequences on the direct victims of domestic abuse and the children who live in the same household. Gaining a deeper understanding of the tangible, far-reaching consequences of domestic abuse will help to design survivor support programmes and quantify the societal harm caused by it.

@@

\textcolor{red}{\textit{Detailed plan}} First, we will explore the causal effect of past experiences of domestic abuse on present socio-economic outcomes. First, using the CSEW, we will explore how witnessing abuse at home in childhood affects educational, employment and health outcomes in adulthood. The 2016 CSEW included a module asking respondents about whether they witnessed domestic abuse at home as a child. Descriptive statistics of this dataset revealed that those who reported to have witnessed abuse as a child were significantly more likely to report long-term health problems, and live in a single parent household as an adult \cite{ONSChildhood}. However, due to the possibility of various confounders (e.g., those growing up in economically deprived areas are more likely to witnessing abuse, but economic deprivation also affects educational outcomes), these descriptive statistics do not reveal the causal effect of witnessing abuse as a child on socio-economic outcomes in adulthood. To overcome this difficulty in quantifying this causal relationship, we will use propensity score matching, a statistical approach that allows for causal inferences to be drawn about the effect of witnessing abuse on educational, employment, and mental and physical health outcomes, and the probability of experiencing domestic abuse in adulthood. In a similar vein, using propensity score matching and the CSEW, we will explore how past experiences of domestic abuse affect current health outcomes, by comparing people who have experienced domestic abuse in the past, but not in the last 12 months, with people who have never experienced domestic abuse, but otherwise have similar socio-economic characteristics.

Second, using both the CSEW and WMP, we will explore how children's behavioural problems are affected by witnessing domestic abuse at home. Using data from the 10-15 year old questionnaire of the CSEW (which can be linked to the adult questionnaire), we can examine the effect of domestic abuse on children living in such households. The children's questionnaire records information about experiences with bullying, carrying knives, gang membership, school truancy, learning difficulties and health outcomes, including drug use and drinking behaviour. Using this dataset and a propensity score matching approach, we will explore the causal effect of living in an abusive household on childhood behavioural outcomes.

We will complement this analysis using data on young offenders from the WMP. We will identify if young offenders' home addresses can be linked with a domestic abuse incident in the recent past, allowing us to explore whether living in a household with a domestic abuse problem predict the type of crime they will engage in (property-related, public order offence, violent), controlling for the socio-economic characteristics of the neighbourhood (using the IMD). We are especially interested in exploring whether living in an abusive environment is predictive of knife crime-related offences.

@@

\textcolor{red}{\textit{Short intro about aims, highlighting added value.}}
In this research project, we will investigate some of the environmental predictors of domestic abuse. While individual-level characteristics are likely to be the most important predictors of domestic abuse victimisation, the environment around the victim can play a key role in either exacerbating, or alleviating domestic abuse. Understanding how neighbourhood-level characteristics affect the prevalence of domestic abuse and the willingness to report is crucial for the most effective distribution of resources to help victims and targeted awareness campaigns. We will also explore how time-varying, exogenous factors related to alcohol consumption and financial stability affect domestic abuse.

@@

\textcolor{red}{\textit{Detailed plan}} Using the CSEW, we will first explore whether neighbourhood characteristics are predictive of the prevalence of domestic abuse and the willingness to report. Previous literature has suggested various pathways, including reduced collective efficacy of neighbourhood, and normalisation and acceptance of violence through social norms, to explain the link between neighbourhood characteristics and domestic abuse, but the question has not been explored in a UK context yet. First, using a logistic regression approach, we will explore how the interviewer's perception of the street (signs of rubbish, vandalism, general condition of houses, Neighbourhood Watch), as well as the respondent's connection and attitude towards the neighbourhood (length of time living in the local area, noisy neighbours, rubbish lying around, teenagers hanging around, vandalism, drunken and anti-social behaviour, drug trafficking, abandoned cars, speeding traffic, police presence in the area, worries about crime levels) predict domestic abuse victimisation, after controlling for the socio-economic characteristics of the respondent. Second, using a spatial regression framework, we will explore how deprivation in a neighbourhood (LSOA level, measured by the IMD) affects the prevalence of domestic abuse (estimated from the CSEW), and the willingness to report. Using this statistical approach will alleviate the problem of spatial dependency in our dataset, and will allow us to derive a precise estimate of the effect. These analyses will extend our understanding of the neighbourhood-level predictors of domestic abuse victimisation, and complement our individual-level approach outlined in the first project to provide a comprehensive exploration of the factors affecting domestic abuse victimisation and willingness to report.

We are also interested in exploring how time-varying external factors, including alcohol consumption, and financial stress (measured by gambling expenditure and benefit dependency) affect the propensity to engage in abusive behaviours. In investigating this question, we will use the transaction dataset described earlier, which allows us to estimate a spending profile for each LSOA over time. This way we can identify time-specific changes in alcohol consumption, gambling expenses and benefit receipts in parts of the West Midlands, and investigate their effect on the number of reported domestic abuse cases in the area (from the WMP), using a spatial Poisson regression approach.

To explore the effect of alcohol consumption on domestic abuse, we will identify externals events that may affect alcohol consumption (e.g., local festivals, weather, bank holidays, sport tournaments) measured by spending on alcohol in an LSOA in a given time period, and investigate whether it affects the number of reported domestic abuse incidents in that area. Previously, we have found a 60\% increase in alcohol-related domestic abuse when the England national football team won, highlighting the profound effect exogenous events can have on the propensity for violence.

>>>>

 \textcolor{red}{\textit{Short intro about aims, highlighting added value.}} In this project, we will explore the victim and offender characteristics that predict serious harm and the factors determining the decision to report the abuse. Identifying high-risk victim-offender pairs is key in preventing serious harm. Insights from these analyses can inform decisions about the optimal timing and target audience for domestic abuse awareness campaigns.
%
%\textcolor{red}{\textit{What we know so far}}
%
%\href{https://www.thehotline.org/2018/09/28/escalation/}{Escalation} often happens when the abuser feels like they are losing control (e.g., because the survivor threatens to leave); separation is an important risk factors of serious harm; contact over children
%
%\href{http://www.safelives.org.uk/policy-evidence/about-domestic-abuse/how-long-do-people-live-domestic-abuse-and-when-do-they-get}{Reporting}
%
%decision to report: serious violence; friends, family, neighbours notice; happens in public, in front of people

\textcolor{red}{\textit{Detailed plan}} Previous studies investigating risk factors for serious harm mostly relied on police data and focused on identifying high-risk offenders \cite{thornton}. Using a multinomial logistic approach, the CSEW data allows us to investigate the effects of financial independence and resilience (income bracket, employment stability), mental health (self-reported well-being), drug and alcohol dependence (frequency of usage), social isolation (self-reported frequency of going out, living alone or with the perpetrator, family disputes), recent separation, and feeling frightened (increased home security, reporting feeling scared). In the CSEW the level of harm is identified by victims indicating that they required medical attention, or time off work. \TM{maybe delete: "While we cannot investigate causal relationships due to the cross-sectional nature of the CSEW, this analysis will nevertheless provide us with important insights about the characteristics of groups most vulnerable to domestic abuse."}

Using the WMP data, we will also explore the risk factors associated with the perpetrator. The strength of the WMP data is that it has a temporal dimension, and we can follow the same perpetrator-victim pair over time. We will examine whether acts of serous harm can be predicted by previous self-harm and suicide attempts (which required police attendance), previous violent and drug offences, breaches of court orders, co-habitation with and separation from the victim.

@@

\TM{This feels out of place. Should the mis-recording and underreporting be in the same section? Maybe it would make more sense having analyses 3&4 from this section actual put in section 1? If not, then we need to make it clearer what is the distinction.}

We will also use both datasets to explore the predictors of the decision to report and escape the abuse. We are not aware of previous research exploring this question in depth. Using the CSEW and logistic regression approach, we will investigate how the decision to report (or seek external help) depends on the individual-level, time-invariant characteristics of the survivor. Potential factors are trust in the justice system, social ties (living with family, time living in the area, member of Neighbourhood Watch), financial independence (own income, employment status, car ownership), number of children, mental health, and alcohol and drug dependency. The WMP data complements this with time-varying factors. We will use this to examine what are the potential characteristics and triggers that enable a victim to report ongoing, long-term abuse. We are interested in whether these reports are more likely to occur after certain days of the year (e.g., birthday of the victim or perpetrator, Christmas, New Years, Valentines etc.). This is particularly important knowledge for police, so that they can efficiently target awareness campaigns to encourage reporting.

@@

 \textcolor{red}{\textit{Short intro about aims, highlighting added value.}}
 Domestic abuse has long-lasting adverse effects on victims and those close to them. In this research project, our is aim to explore some of these consequences on the direct victims of domestic abuse and the children who live in the same household.
 %Gaining a deeper understanding of the tangible, far-reaching consequences of domestic abuse will help to design survivor support programmes and quantify the societal harm caused by it.

@@

\textcolor{red}{\textit{Detailed plan}} First, we will consolidate and extend research on the effect of witnessing domestic abuse during childhood. The 2016 CSEW included a module asking respondents about whether they witnessed domestic abuse at home as a child. Descriptive statistics of this dataset revealed that those who reported to have witnessed abuse as a child were significantly more likely to report long-term health problems, and live in a single parent household as an adult \cite{ONSChildhood}. We will extend this using propensity score matching, a statistical approach that allows for causal inferences to be drawn about the effect of witnessing abuse whilst also controlling for a range of confounds, such as economic deprivation. \TM{Do we have any info on economic deprivation of where they grew up, or just where they live now?}

Second, we will explore how children's behavioural problems are affected by witnessing domestic abuse at home. The CSEW questionnaire for 10-15 year olds records information about experiences with bullying, carrying knives, gang membership, school truancy, learning difficulties and health outcomes, including drug use and drinking behaviour. By applying propensity score matching, we will explore the causal effect of living in an abusive household.

We will complement this analysis using data on young offenders from the WMP. We will identify young offenders whose home addresses has been linked with a domestic abuse incident in the recent past. It is not possible to predict whether witnessing domestic abuse increases the chance of becoming a young offender, but we can examine whether the type of crime they engage in. For example, whether, after controlling for controlling for the socio-economic characteristics of the neighborhood (using the IMD), such individuals are more likely to commit similarly violent acts than other young offenders.

@@

\textcolor{red}{\textit{Short intro about aims, highlighting added value.}}
In this research project, we will investigate some of the environmental predictors of domestic abuse. While individual-level characteristics are likely to be the most important predictors of domestic abuse victimisation, the environment around the victim can play a key role in either exacerbating, or alleviating domestic abuse.

@@

\textcolor{red}{\textit{Detailed plan}} The CSEW includes rich information about the individual's environment, including the interviewer's perception of the street (signs of rubbish, vandalism, general condition of houses etc.), as well as the respondent's connection and attitude towards the neighbourhood (time living in the local area, noisy neighbours, teenagers hanging around, vandalism, drunken and anti-social behaviour, drug trafficking, abandoned cars, police presence in the area etc.). These properties of the environment can be used to predict domestic abuse victimisation, whilst controlling for the socio-economic characteristics of the respondent. Additionally, using a spatial regression framework, we will explore how deprivation in a neighbourhood (LSOA level, measured by the IMD) affects the prevalence of domestic abuse and the willingness to report. Using this statistical approach will alleviate the problem of spatial dependency in our dataset, and will allow us to derive a precise estimate of the effect. These analyses will extend our understanding of the neighbourhood-level predictors of domestic abuse victimisation, and complement our individual-level approach outlined in the first project to provide a comprehensive exploration of the factors affecting domestic abuse victimisation and willingness to report.

We are also interested in exploring the effect of factors including average alcohol consumption, and levels of financial stress at the community level, rather than the individual (measured by gambling expenditure and benefit dependency). In investigating this question, we will use the transaction dataset described earlier, which allows us to estimate a spending profile for each LSOA over time. This way we can identify time-specific changes in alcohol consumption, gambling expenses and benefit receipts in parts of the West Midlands, and investigate their effect on the number of reported domestic abuse cases in the area (from the WMP), using a spatial Poisson regression approach. \TM{Can we not do this for the whole country using the CSEW? I know the timecourse isn't great, but we have the data from individual years.}

We shall complement this analysis by identifying external events that may affect alcohol consumption (e.g., festivals, weather, bank holidays, sport tournaments) measured by spending on alcohol in an LSOA in a given time period, and investigate whether it affects the number of reported domestic abuse incidents in that area. Previously, our research has identified a 60\% increase in alcohol-related domestic abuse when the England national football team won, highlighting the profound effect exogenous events can have on the propensity for violence.
